---
title: "Intro to coding with R"
---

This is an R Notebook. When you execute code within the notebook, the results appear beneath the code.

Thinking of our R session as an empty factory, we first need to get some machines (functions). We'll install two packages, also called libraries, of functions that will be useful to us. Run the code below (by pressing the green *play* arrow at the top right of the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*) to install the `tidyverse` and `immunogenetr` libraries.

```{r}
install.packages("tidyverse")
install.packages("devtools")

library(devtools)

install_github("k96nb01/immunogenetr_package")
```

Great, we have our packages of useful functions installed. There's one more step before we can use them - we have to "load" them into our R factory. Run the code chunk below to load the `tidyverse` and `immunogenetr` packages.

```{r}
library(tidyverse)
library(immunogenetr)
```
Above we should see an indication that the `tidyverse` package was loaded (the `immunogenetr` package doesn't tell us anything, but it loaded). This gives us lots of great functions to work with. Let's use one of these functions to import some data into our R factory. Run the code chunk below:

```{r}
read_csv("HLA_haplotypes.csv")
```

We've just run our first line of code in R! To explain what we just did - we used a function (machine) called `read_csv`. Immediately after the name of the function (`read_csv`), we put in the arguments (settings) for the function. The arguments for this function were the name of the file we wanted to read ("HLA_haplotypes.csv"). The program we're using showed us the output of the function in the space above. The first thing it output was some information about what it did to read the function (click on the small box that says "R Console" above). The second thing it output (click on the box labeled "spec_tbl_df" above) is the imported data in a format called a tibble, a type of data frame. Take a moment to look at the data. Try clicking on the small right arrow at the top right of the data frame. What can you learn about this data frame?



OK, so we've run our first function in R, which read an HLA haplotype table. However, what we did was essentially drop the data into the middle of the machine. We want to set up an assembly line for our data, though, so let's instead feed our data into the `read_csv` function using what's known as a "pipe." This piece of code: ` %>% ` is known as a pipe, and it functions like a track in a factory. Let's try running the code again, this time using the pipe:

```{r}
"HLA_haplotypes.csv" %>% read_csv()
```

The output, above, should look exactly the same as in the first code chunk. However, we've set up our factory assembly line (often referred to as a pipeline, and since we're using a pipe, ` %>% ` we now know why!). Let's extend our assembly line by piping this to another function: `select`. Extend the pipeline by adding this to the end of the code: `%>% select()`

```{r}
"HLA_haplotypes.csv" %>% read_csv()
```

Uh-oh, what happened to our data? It looks like we now have an empty data frame. Perhaps we didn't understand how to use the function, and which arguments to give the function. Let's see if we can learn a bit about the function by bringing up it's help file. The easiest way to do this is to type "?" and then the function into the console,  all the way in the bottom left of the browser window, where you see ">". Go ahead and type this into the console "?select()".


The help file for the function is really helpful! It tells us what the function does, how to use the arguments, and even gives lots of examples of how to use the function. It looks like this function will take the data frame and select the columns we want. Let's try giving an argument to the function so that just the DRB1 column is selected. Let us know if you need help figuring out how to do this.

```{r}
# Add an argument to this function to select the DRB1 locus:

"HLA_haplotypes.csv" %>% read_csv() %>% select()
```

Awesome, by setting the function up like this: `select(DRB1)`, we subsetted our data frame to only contain the DRB1 column! Our data pipline is starting to look like a sentence, but if we keep adding things at the end it could get hard to read. Luckily, we can put each function on a new line, which often makes things easier to read. Take a look at the code below, and try running it:

```{r}
"HLA_haplotypes.csv" %>% 
  read_csv() %>% 
  select(DRB1)
```

Now we have each function on a new line, which makes it easier to see what is happening to our data at each step. We'll keep adding things to our pipeline in new lines.

Let's extend our pipeline with another function: `filter`. The `filter` function will keep any row in our data frame that meets the criteria we set up in the function's arguments. Let's try filtering for DRB1*11:01:01:01.

```{r}
# Set up the arguments of the `filter` function so that we filter for "DRB1*11:01:01:01". Here's a hint: in R, saying something equals something else requires two equal signs (==), and make sure you put "DRB1*11:01:01:01" in quotes! 

"HLA_haplotypes.csv" %>% 
  read_csv() %>% 
  select(DRB1) %>% 
  filter()
```

OK, we now have filtered for DRB1*11:01:01:01. The next thing we'll try is a very useful function called `mutate`. The `mutate` function makes a new column in data frames. Try running the code below and look what happens.

```{r}
"HLA_haplotypes.csv" %>% 
  read_csv() %>% 
  select(DRB1) %>% 
  filter(DRB1 == "DRB1*11:01:01:01") %>% 
  mutate(DRB1_copy = DRB1)
```

We can see that the `mutate` function created a new column, and in the arguments we named that new column "DRB1_copy", and then filled that column with values from the "DRB1" column. That's great, but the real power of the `mutate` function is the ability to compute new data, not just copy other data. To do that, we can use a function inside of the `mutate` function (functions within functions - it's like Inception!). Let's say we want to truncate the 4-field typing in the DRB1 column down to two fields, and put that into a new column. How might we go about doing that? There are lots of ways, but luckily, the `immunogenetr` package has a function called `HLA_truncate` that will do exactly what we want. Try running the code below and seeing what happens:

```{r}
"HLA_haplotypes.csv" %>% 
  read_csv() %>% 
  select(DRB1) %>% 
  filter(DRB1 == "DRB1*11:01:01:01") %>% 
  mutate(DRB1_truncated = HLA_truncate(DRB1))
```

What did the `HLA_truncate` function do? If you have time, can you figure out how to have the function truncate to 3 fields instead of 2?

OK, we've set up a nice pipeline, and we'd like to stop for now. The way we have the pipeline set up now, our data goes to the `mutate` function, but then stops there. Using our factory metaphor, our data is coming out of the last function and falling on the floor. Let's instead collect it in a bin. We do this in R by assigning it to an "object." To assign our data to an object we can use the assignment operator `<-`. You'll note that this looks like an arrow pointing the wrong way, and I admit, it is a bit jarring, but to assign something to an object, we have to put it at the very beginning of the pipeline, as so:

```{r}
# The code below adds `HLA_object <-` at the beginning. Run the code chunk to see what happens.

HLA_object <- "HLA_haplotypes.csv" %>% 
  read_csv() %>% 
  select(DRB1) %>% 
  filter(DRB1 == "DRB1*11:01:01:01") %>% 
  mutate(DRB1_truncated = HLA_truncate(DRB1))
```

Wait, where did our data go? All we see now is the information about the `read_csv` function! What happened is that in all the code we ran above, since we were never assigning our output to anything, it showed us the results just to prove that it did something. Now that we've assigned it to an object we created called `HLA_object`, we've told R to do something, and so it doesn't show us what's in the data we created. If we want to see what's in our object, we can use the `print` function (note that in computer coding, `print` should be though of as "show me", rather than "print this out on paper"). Run the code below to look at the data we created.

```{r}
print(HLA_object)
```

Having an object in R is really useful. We can use it in anything else in R, even inside of functions, or use it at the start of another pipeline. Let's say we're done for now, though, and instead export the data. In the code below we'll send our object to the `write_csv` function to save a file with our data:

```{r}
HLA_object %>% write_csv("HLA_object.csv")
```

You should see that you have a file called "HLA_object.csv" in your Files pane to the right.

And there you go, you've performed a full set of data manipulation on some HLA data. Using our factory metaphor, we've trucked some data in our R factory, set up several machines linked by a track, collected the results in a bin, and then shipped our finished data out. Understanding and using these base principles, you can do really some really interesting informatics work in HLA!
